\name: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-03-04 paper/main.tex
.

Within DTI prediction, there are potential biases resulting from the
underlying datasets \citep{Pahikkala2014}. First, novel drugs are
often designed by altering non-functional components of a drug,
leading to two and more very similar drugs designed to target the same
proteins \cite{Overington2006}. This can result in a bias when it
leads to hidden duplicates that can distribute among the train/test
split, resulting in a better (measured) predictive performance than
would be expected when the model is applied to identify drugs that
target a protein for which no drugs yet exist. Second, some proteins
(which we will denote as \textit{hub proteins}) have significantly
more known interactions with drugs than others. In the STITCH
database, $5\%$ of the proteins have $40\%$ of the interactions, and
similar distributions are present in the Yamanishi and Drugbank
datasets \cite{Drugbank2007, Drugbank2017} datasets; preferentially
predicting these proteins may increase predictive performance while
again not reflecting the actual performance when applied to a new
protein (e.g., a protein for which no interactions are known). These
differences in the number of drugs targeting certain proteins may be
the result of study bias where more ``valuable'' proteins have more
drugs designed to target them due to their involvement in more common
diseases (or diseases for which drugs can be more profitably
marketed).

We developed \name as a computational model to predict drug--target
interactions. Specifically, given a protein, \name will identify and
rank known drugs that likely target this protein. \name combines two
types of features: structural information for drugs and proteins that
can be used to determine if they may physically interact, and
information about drug effects that may localize on an interaction
network.  As structural features, \name uses structural
representations of drugs from the SMILES transformer
\cite{SmilesTransformer} and representations of protein amino acid
sequences from DeepGOPlus \cite{DeepGoPlus}.  \name learns
representations of drug effects and protein functions using the
ontology-based machine learning method DL2Vec \cite{DL2vec2020}.

@@

We use two Siamese neural networks to combine the molecular and
phenotype representations of drugs and proteins, and add these to a
protein interaction network as node labels.  We then train this model
in an end-to-end manner using a graph neural network architecture.
The overall workflow\todo{make sure it's a workflow} is depicted in
Figure~\ref{fig:Fullmodel}.

>>>>

\todo[inline]{Introduction}
Within DTI prediction, there are potential biases resulting from the
underlying datasets \citep{Pahikkala2014}. First, novel drugs are
often designed by altering non-functional components of a drug,
leading to two and more very similar drugs designed to target the same
proteins \cite{Overington2006}. This can result in a bias when it
leads to hidden duplicates that can distribute among the train/test
split, resulting in a better (measured) predictive performance than
would be expected when the model is applied to identify drugs that
target a protein for which no drugs yet exist. Second, some proteins
(which we call \textit{hub proteins}) have significantly more known
interactions with drugs than others. In the STITCH database, $5\%$ of
the proteins have $40\%$ of the interactions, and similar
distributions are present in the Yamanishi and Drugbank
\cite{Drugbank2007, Drugbank2017} datasets; preferentially predicting
these proteins may increase predictive performance while not
reflecting the actual performance when applied to a new protein (e.g.,
a protein for which no interactions are known). These differences in
the number of drugs targeting certain proteins may be the result of
study bias where more ``valuable'' proteins have more drugs designed
to target them due to their involvement in more common diseases (or
diseases for which drugs can be more profitably marketed).

We developed \name as a computational model to predict drug--target
interactions. Specifically, given a protein, \name will identify and
rank drugs that likely target this protein. \name combines two types
of features: structural information for drugs and proteins that can be
used to determine if the drug and protein physically interact, and
information about phenotypic effects of drugs and changes in protein
function that may ``localize'' on an interaction network (i.e.,
neighboring nodes will share some of these features or be
phenotypically similar).  As structural features, \name uses
structural representations of drugs from the SMILES transformer
\cite{SmilesTransformer} and representations of protein amino acid
sequences from DeepGOPlus \cite{DeepGoPlus}.  \name learns
representations of drug effects and protein functions using the
ontology-based machine learning method DL2Vec \cite{DL2vec2020} and
ontology-based annotations of drugs and proteins.

@@

We use two Siamese neural networks to combine the molecular and
phenotype representations of drugs and proteins. Siamese neural
networks are used to learn a kind of similarity between the molecular
and phenotype representations. We add these similarities to a protein
interaction network as node labels; for each drug, proteins receive
two features (from the molecular and phenotype similarities of the
drug to the protein).
\todo[inline]{There is a missing sentence here explaining the graph
  neural network, output, etc.}
We then train this model in an end-to-end manner using a graph neural
network architecture.  The overall workflow\todo{make sure it's a
  workflow} is depicted in Figure~\ref{fig:Fullmodel}.
